An Integral Projection Modeling Approach to Understanding Demographic Effects of Multispecies Mutualisms

### 2024-01-19 Manuscript/ant_cactus_AmNat.tex
thursday

\section*{Abstract}
Mutualisms are widespread species interactions with diverse and dynamic consequences. 
They are considered more context dependent than other species interactions, meaning there are many different factors which change the outcomes of interactions between mutualists, including partner diversity. 
Partner diversity has become a central focus in the field of mutualisms, expanding previous work from primarily pairwise to multispecies mutualisms. 
It has been shown that pairwise studies are poor predictors of the effects of multispecies mutualistic interactions. 
The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.

@@

Mutualism is defined at the level of a species pair (+/+) but these interactions are embedded within multi-species communities, and growing evidence suggests that pairwise interactions are poor predictors of the net effects of multi-species mutualism \citep{Afkhami2014,Palmer2010,Bascompte2009,Dattilo2014}. 
A focal mutualist may interact with multiple guilds of partner types (e.g., plants that interact with pollinators, seed dispersers, soil microbes, and ant defenders) or with multiple partner species within the same guild (e.g., plants visited by multiple pollinator species). 
Within a mutualist guild, partner species often differ in the amount or type of goods or services they provide, making partner identity an important source of contingency in mutualism \citep{Stanton2003}. 
Whether and how partner diversity modifies the demographic effects of mutualistic interactions remain open questions within relevance in applied settings \citep{rogers2014}.

There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \cite{Yeung2006,Barrett2015,Ushio2020}. 
First, when there is a hierarchy of fitness effects (a consistent ranking of best to worst mutualists), a more diverse sample of the partner community may be more likely to include the best partner \cite{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \cite{Batstone2018}. 
However, if partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
Second, even within a single mutualist guild, the benefits conferred by alternative partner species can vary in type, and not just degree \cite{Stachowicz2005,Bronstein2006,Stanton2003}. 
This can lead to a positive effect of partner diversity through complementarity of alternative functions \cite{Batstone2018}. 
Interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \cite{Afkhami2014}. 
Third, partner species and herbivores can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits and protection across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \cite{Batstone2018,Lazaro2022,Horvitz1990}.

@@

Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006}.
Presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014}, and visitation by low quality ant partners can prevent visitation by higher quality partners, consequently causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the effectiveness of ant partners.
Temporal dynamics of ant visitation therefore can have important impacts on the fitness of the plant partners \citep{Barton2010, Boege2005, Fonseca2003}.
Herbivore identity and pressure can vary inter-annually \cite{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face varies just as much as the protection they recieve due to temporal stochasticity. 
Recent studies have begun to investigate how ant partner diversity affects plant fitness \citep{Palmer2010,Afkhami2014,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}
However, little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because the long-term data required to detect portfolio effects are rarely available.

This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindriopuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Previous studies have shown that herbivory by specialized insect herbivores negatively affects plant fitness \cite{Miller2009}, and ant defense reduces herbivore damage \cite{Miller2007}. 
Tree cholla are tended by two common ant species (\textit{Liometopum apiculatum} and \textit{Crematogaster opuntiae}) and several infrequent species, all of which are ground-nesting. 
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{Liometopum apiculatum} providing strong fitness benefits and \textit{Crematogaster opuntiae} having net negative fitness effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}. 
However, those studies did not integrate the demographic effects of ant defense across the plant life cycle, nor did they account for inter-annual fluctuations in the herbivore populations.

We used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species and how the demographic effects of alternative partner species varied across plant size structure and nearly 20 years of inter-annual fluctuations. 
We used this observational data set, contextualized by previous experiments, to ask whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation for the focal plant partner. 
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the demographic effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
We used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) that combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.

@@

This study was conducted in the Los Pi$\tilde{n}$os mountains, a small mountain chain located on the Sevilleta National Wildlife Refuge, a Long-term Ecological Research site (SEV-LTER) in central New Mexico, USA.
This is an area characterized by steep, rocky slopes, and perennial vegetation including grasses (\textit{Bouteloua eriopoda} and \textit{B. gracilis}), yuccas, cacti, and junipers. 
Tree cholla cacti are common in high Chihuahuan desert habitats, with their native range spanning the southwestern USA \citep{Benson1982}. 
These arborescent plants produce cylindrical segments with large spines. 
In the growing season (May to August in New Mexico), the plants initiate new vegetative segments and flower buds at the ends of existing segments. 
While most plants produce new segments every season, only those which are reproductively mature produce flower buds. 
%Tree cholla generally reach at least 9 years of age before beginning to reproduce \citep{Ohm2014}.
Like other EFN-bearing cacti, tree cholla secrete nectar from specialized glands on young vegetative segments and flower buds \citep{Ness2006,Oliveira1999}. 
Flower buds produce more and higher-quality EFN than vegetative segments, making reproductive cholla valuable mutualist partners.

Tree cholla EFN is harvested by various ant species. 
At SEV-LTER, cholla are visited primarily by two species of formicoid, ground-nesting ants, \textit{Crematogaster opuntiae} and \textit{Liometopum apiculatum}, as well as other rarer species, including \textit{Forelius pruinosus} and unidentified species of \textit{Aphaenogaster} and \textit{Camponotus}.
\textit{L. apiculatum} are the most frequent visitors with $25\% - 60\%$ of tree cholla tended by these ants, followed by \textit{C. opuntiae} visiting between $5\% - 20\%$ of cacti \citep{Donald2022} depending on the year. Between $ 30\% 80\%$ of cacti remain vacant in any given year. 
These ants rarely co-occur on a plant, likely due to interspecific competition \citep{Miller2007}: staged introductions of \textit{C. opuntiae} to \textit{L. apiculatum}-tended plants, and vice versa, provoke aggressive responses by resident ants (A. Cambpell, \textit{personal observation}).
Each cholla is visited by a single ant species for the duration of a season, and the species of the visitors can change from one season to the next. 
At the beginning of the growing season, when EFN production begins, the ground-nesting ants will begin visiting tree cholla.
They will visit the cholla every day during the season around the clock, with the most acticity around sunrise or sunset \citep{Ohm2014}. 
Smaller cholla are less likely to be visited because they produce very little EFN, so larger cholla, especially flowering individuals, are generally more highly tended \citep{Miller2014}. 
In late August, the tree cholla stop producing EFN and the ants vacate until the next growing season.

There are a variety of insect herbivores and seed predators which attack the cholla \citep{Mann1969}. 
An unidentified weevil of the genus \textit{Gerstaekeria} feeds on vegetative and reproductive structures and implants their larvae within the plant tissue for the winter. 
A cactus bug, \textit{Narnia pallidicornis}, (Hemiptera: Coreidae) feeds on all cholla parts with a preference for the reproductive structures \citep{Miller2006}.
A seed predator, \textit{Cahela ponderosella}, (Lepidoptera: Pyralidae) attacks developing fruits pre-dispersal and oviposits in open flowers mid-growing season where larvae burrow into the ripening ovary. 
These predators can have significant negative impacts on the fitness of individual cholla and depress population growth \citep{Miller2009}.
There is experimental evidence that tree cholla tended by \textit{L. apiculatum} and \textit{C. opuntiae} experience less herbivory than plants from which ants were excluded \citep{Miller2007}.

\subsection*{Data Collection}
This study is based a long-term demographic data set spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the covid shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded the height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We recorded reproductive effort as counts of viable and aborted flowerbuds. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (QUANTITY), and in these cases we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
In total, the data set included \# unique individuals and \# plant-year observations. 
These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

@@

Given the low frequency of ant species other than \textit{L. apiculatum} and \textit{C. opuntiae} (QUANTIFY THIS) we combined observations of all other ants into an ``other'' category, such that our models included four possible ant states: vacant, \textit{L. apiculatum}, \textit{C. opuntiae}, and ``other''. 
The ``other'' category was made of unidentified ant species and ant species which occurred at relatively low frequencies compared to \textit{C. opuntiae} and \textit{L. apiculatum} (such as unknown species belonging to the genus \textit{Aphenogaster} and unknown species belonging to the genus \textit{Camponotus}).
Ant state is included as a predictor variable in sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
Ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, and flowerbud viability. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering and flowerbud production, independent of its influence on plant size. 
Therefore, these sub-models included plant size but not ant state as predictor variables.

@@

Functions $P(x';\pmb{\tau^P})$ and $F(x';\pmb{\tau^F})$ give the probability of flowering and the number of flowerbuds produced, respectively, by plants of size $x'$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau^V_{a}})$) and the number of seeds per fruit ($\kappa(a')$) are dependent on ant state $a$ but not size. 
The vector $\pmb{\tau}$ gives year-specific deviates (with mean zero) that modify the intercepts of the functions in which it appears. 
This vector appears in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that deviates are specific to plants in ant state $a$. 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

For the above-ground part of the life cycle, the number of plants of size $x'$ and ant state $a'$ in year $t+1$ ($n(x',a')_{t+1}$) is given by survival/growth transitions from size $x$ and ant state $a$ in year $t$, plus germination out of the seed banks:
\begin{linenomath*}
	$$
	n(x',a')_{t+1} = (\gamma_1 B^1_{t} + \gamma_2 B^2_{t} ) \eta(x') \omega \rho(a')  + \\
	$$
	$$
	\sum_{a=1}^{A} \int_L^U S(x,a;\pmb{\tau^S_{a}}) G(x',x,a;\pmb{\tau^G_{a}}) \epsilon(x,a,a';\pmb{\tau^{\epsilon}}) n(x,a)_t dx \\
	$$
\end{linenomath*}

The first term estimates the number of individuals recruiting from a one or two-year seed bank to a plant of size $x'$ and ant state $a'$ based on the recruit size distribution $\eta(x')$ and the proportion of seedlings which survive from germination (late summer) to the census (May) $\omega$.
This term is multiplied by $\rho(a')$, which gives the probability that a new recruit has ant state $a'$ at its first appears in our census ($\sum_{a'}^{A}\rho(a')=1$). 
The second term represents all possible transitions from size $x$ and ant $a$ to size $x'$ and ant $a'$, conditioned on survival. 
Survival ($S(x,a;\pmb{\tau^S_{a}})$) and growth ($G(x',x,a;\pmb{\tau^G_{a}})$) are both dependent on initial size and ant state. 
As above, these functions accommodate inter-annual variability through year-specific intercepts that can vary by ant state ($\pmb{\tau_{a}}$). 
Ant transition function $\epsilon(a',a,x;\pmb{\tau^{\epsilon}})$ gives the probability that a plant with size $x$ and ant state $a$ transitions to ant state $a'$ in the next census. 
This function accomodates inter-annual variability through year-specific intercepts which are stable across ant states ($\pmb{\tau}$).

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models (GLMMs) in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transition between ant states conditional on plant size and previous ant state. 
All models were fit using Stan and Rstan (CITE). 
Unless otherwise mentioned, all models use vague priors.

@@

\paragraph{Seeds Per Flower}
With data from \tom{Miller 2007}{CITE}********, we fit a Bayesian generalized linear model for the number of seeds produced by every flower on a cholla ($\kappa(a')$) in year $t+1$ based on the ant partner $a'$ in year $t+1$, using a Negative Binomial distribution:
$$y^{\kappa} \sim 0 Truncated Negative Binomial(\hat{\kappa},\hat{\phi})$$
$$ \hat{\kappa } = \beta_{0}^{\kappa} \times a'$$
$$\hat{\phi} = \beta_{0}^{\phi}$$
Ant partners are included as predictors here because they reduce floral abortion rates and therefore may lead to higher numbers of seeds.

>>>>

\section*{Abstract}
\tom{}{I think this is too long for Am Nat requirements. Also, use ``we''.}
Mutualisms are widespread species interactions with diverse and dynamic consequences. 
They are considered more context dependent than other species interactions, meaning there are many different factors which change the outcomes of interactions between mutualists, including partner diversity. 
Partner diversity has become a central focus in the field of mutualisms, expanding previous work from primarily pairwise to multispecies mutualisms. 
It has been shown that pairwise studies are poor predictors of the effects of multispecies mutualistic interactions. 
The diversity of partners in a multi-species mutualism causes varied demographic effects on the population of the focal mutualist which can be explained by several mechanisms: portfolio effect, complementarity, and sampling effect.

@@

Mutualism is defined at the level of a species pair (+/+) but these interactions are embedded within multi-species communities, and growing evidence suggests that pairwise interactions are poor predictors of the net effects of multi-species mutualism \citep{Afkhami2014,Palmer2010,Bascompte2009,Dattilo2014}. 
A focal mutualist may interact with multiple guilds of partner types (e.g., plants that interact with pollinators, seed dispersers, soil microbes, and ant defenders) or with multiple partner species within the same guild (e.g., plants visited by multiple pollinator species). 
Within a mutualist guild, partner species often differ in the amount or type of goods or services they provide, making partner identity an important source of context-dependence in mutualism \citep{Stanton2003}. 
Whether and how partner diversity modifies the demographic effects of mutualistic interactions remain open questions within relevance in applied settings \citep{rogers2014}.

There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \cite{Yeung2006,Barrett2015,Ushio2020}. 
First, when there is a hierarchy of fitness effects (a consistent ranking of best to worst mutualists), a more diverse sample of the partner community may be more likely to include the best partner \cite{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \cite{Batstone2018}. 
However, if partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
Second, even within a single mutualist guild, the benefits conferred by alternative partner species can vary in type and not just degree \cite{Stachowicz2005,Bronstein2006,Stanton2003}. 
This can lead to a positive effect of partner diversity through complementarity of alternative functions \cite{Batstone2018}. 
Interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \cite{Afkhami2014}. 
Third, partner species and \tom{herbivores}{notice that ``herbiviores'' does not fit here, at least not yet. you have not said anything about ants and plants, you are talking very generally. put yourself in the mind of a reader as you write, this will help ensure that ideas appear in an order that can be digested.} can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits and protection across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \cite{Batstone2018,Lazaro2022,Horvitz1990}.

@@

Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
Presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014}, and visitation by low quality ant partners can prevent visitation by higher quality partners, consequently causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Herbivore identity and pressure can vary inter-annually \cite{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
Recent studies have begun to investigate how ant partner diversity affects plant fitness \citep{Palmer2010,Afkhami2014,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}
However, little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because the necessary long-term data are rarely available.

This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindriopuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Previous studies have shown that herbivory by specialized insect herbivores negatively affects plant fitness \cite{Miller2009}, and ant defense reduces herbivore damage \cite{Miller2007}. 
Tree cholla are tended by two common ant species (\textit{Liometopum apiculatum} and \textit{Crematogaster opuntiae}) and several infrequent species, all of which are ground-nesting. 
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{Liometopum apiculatum} providing strong fitness benefits through anti-herbivore defense and \textit{Crematogaster opuntiae} having net negative fitness effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}. 
However, those studies did not integrate the demographic effects of ant defense across the plant life cycle, nor did they account for inter-annual fluctuations in the \tom{herbivore populations}{I would be careful here -- unless you are using the herbivore counts I don't think you have data on this.}.

We used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal turnover in partner identity at the individual level, and how the demographic effects of alternative partner species varied across plant size structure and nearly 20 years of inter-annual fluctuations. 
We used this observational data set, contextualized by previous experiments, to ask whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation for the focal plant partner. 
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the demographic effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
We used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) that combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.

@@

This study was conducted in the Los Pi$\tilde{n}$os mountains, a small mountain chain located on the Sevilleta National Wildlife Refuge, a Long-term Ecological Research site (SEV-LTER) in central New Mexico, USA.
This is an area characterized by steep, rocky slopes, and perennial vegetation including grasses (\textit{Bouteloua eriopoda} and \textit{B. gracilis}), yuccas, cacti, and junipers. 
Tree cholla cacti are common in high Chihuahuan desert habitats, with their native range spanning the southwestern USA \citep{Benson1982}. 
These arborescent plants produce cylindrical segments with large spines. 
In the growing season (May to August in New Mexico), the plants initiate new vegetative segments and flower buds at the ends of existing segments. 
While most plants produce new segments every season, only those which are reproductively mature produce flower buds. 
%Tree cholla generally reach at least 9 years of age before beginning to reproduce \citep{Ohm2014}.
Like other EFN-bearing cacti, tree cholla secrete nectar from specialized glands on young vegetative segments and flower buds \citep{Ness2006,Oliveira1999}. 
Flower buds produce more and higher-quality EFN than vegetative segments, making reproductive cholla valuable mutualist partners. 
Smaller cholla produce little to no EFN, so larger cholla, especially flowering individuals, are generally more highly tended \citep{Miller2014}.

Tree cholla EFN is harvested by various ant species. 
At SEV-LTER, cholla are visited primarily by two species of ground-nesting ants, \textit{Crematogaster opuntiae} and \textit{Liometopum apiculatum}, as well as other rarer species, including \textit{Forelius pruinosus} and unidentified species of \textit{Aphaenogaster} and \textit{Camponotus}.
\textit{L. apiculatum} are the most frequent visitors with $25\% - 60\%$ of tree cholla tended by these ants, followed by \textit{C. opuntiae} visiting between $5\% - 20\%$ of cacti depending on the year \citep{Donald2022}. 
Between $ 30\% 80\%$ of cacti remain vacant in any given year. 
These ants rarely co-occur on a plant, likely due to interspecific competition \citep{Miller2007}: staged introductions of \textit{C. opuntiae} to \textit{L. apiculatum}-tended plants, and vice versa, provoke aggressive responses by resident ants (A. Cambpell, \textit{personal observation}).
Each cholla is visited by a single ant species for the duration of a season, and the species of the visitors can change from one season to the next. 
In late August, the tree cholla stop producing EFN and the ants vacate until the next growing season.

There are a variety of insect herbivores and seed predators which specialize on tree cholla \citep{Mann1969}. 
A weevil of the genus \textit{Gerstaekeria} feeds on vegetative and reproductive structures and implants their larvae within the plant tissue for the winter. 
A cactus bug, \textit{Narnia pallidicornis}, (Hemiptera: Coreidae) feeds on all cholla parts with a preference for the reproductive structures \citep{Miller2006}.
A seed predator, \textit{Cahela ponderosella}, (Lepidoptera: Pyralidae) attacks developing fruits pre-dispersal and oviposits in open flowers mid-growing season where larvae burrow into the ripening ovary. 
These predators can have significant negative impacts on plant fitness of and depress population growth \citep{Miller2009}.
There is experimental evidence that tree cholla tended by \textit{L. apiculatum} and \textit{C. opuntiae} experience less herbivory than plants from which ants were excluded \citep{Miller2007}.

\subsection*{Data Collection}
This study is based a long-term demographic data set spanning 2004 to 2023 at SEV-LTER. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots has since been censused annually through 2023 (with the exception of 2020 due to the covid shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded the height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We recorded reproductive effort as counts of viable and aborted flowerbuds. 
We recorded the ant species present (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (\tom{}{quantify}), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
\tom{In total, the data set included \# unique individuals and \# plant-year observations. }{update numbers}
These data were used to fit vital rate models (survival, growth, reproduction) as functions of plant size and ant occupancy state.

@@

Given the low frequency of ant species other than \textit{L. apiculatum} and \textit{C. opuntiae} (\tom{}{Quantify this.}) we combined observations of all other ants into an ``other'' category, such that our models included four possible ant states: vacant, \textit{L. apiculatum}, \textit{C. opuntiae}, and ``other''. 
\tom{The ``other'' category was made of unidentified ant species and ant species which occurred at relatively low frequencies compared to \textit{C. opuntiae} and \textit{L. apiculatum} (such as unknown species belonging to the genus \textit{Aphenogaster} and unknown species belonging to the genus \textit{Camponotus}).}{I think you can do better than this because we actually have the designated species in the data. I would report here the actual breakdown (percentages) of species in the ``other'' group. I expect Forelious will be a big chunk of it.}
Ant state is included as a predictor variable in sub-models where there are biologically realistic pathways through which ants could impact the outcome of that process. 
For example, ant partners defend cacti from herbivores, and prior experimental work indicates that ant tending can reduce vegetative tissue loss and floral abortion.
Therefore, ant state was included in sub-models for survival, growth, and flowerbud viability. 
In contrast, we have no reason to expect that ant tending can directly influence the probability of flowering and flowerbud production, independently of its influence on plant size. 
Therefore, these sub-models included plant size but not ant state as predictor variables.

@@

\noindent Functions \tom{$P(x';\pmb{\tau^P})$ and $F(x';\pmb{\tau^F})$}{These should be $x$ not $x'$ because we modeled flowering and fertility in year t based on size in year t.} give the probability of flowering and the number of flowerbuds produced, respectively, by plants of size $x'$ in year $t$. 
The proportion of flowerbuds that remain viable through fruit set ($V(a;\pmb{\tau^V_{a}})$) and the number of seeds per fruit (\tom{$\kappa(a')$}{Why is this $a'$? I think it should be $a$. More generally, I think you need to explain your use of primes and how readers should interpret them at the start of model exposition.}) are dependent on ant state $a$ but not size. 
The vector $\pmb{\tau}$ gives year-specific deviates (with mean zero) and appears in functions for which we can estimate temporal stochasticity from the long-term data; superscripts indicate the corresponding vital rate and subscripts indicate that \tom{deviates are specific to plants in ant state $a$}{Why does the tau vector for viability have the $a$ subscript but the others do not?}. 
Seed production is integrated over the size distribution, from the lower $L$ to upper $U$ size limits, and summed over all possible ant states ($A$) giving total seed production. 
Seeds are multiplied by the probability of seed dispersal and survival ($\delta$) to give the number of seeds that enter the one-year old seed bank. 
Plants can recruit out of the year-one seed bank with probability $\gamma_1$ or transition to the two-year seed bank with a probability of $1 - \gamma_1$. 
Seeds in the two-year seed bank are assumed to either germinate with probability $\gamma_2$ or die.

For the above-ground part of the life cycle, the number of plants of size $x'$ and ant state $a'$ in year $t+1$ ($n(x',a')_{t+1}$) is given by survival/growth transitions from size $x$ and ant state $a$ in year $t$, plus germination out of the seed banks:
\begin{linenomath*}
	$$
	n(x',a')_{t+1} = (\gamma_1 B^1_{t} + \gamma_2 B^2_{t} ) \eta(x') \omega \rho_{0}(a')  + \\
	$$
	$$
	\sum_{a=1}^{A} \int_L^U S(x,a;\pmb{\tau^S_{a}}) G(x',x,a;\pmb{\tau^G_{a}}) \rho(x,a,a';\pmb{\tau^{\epsilon}}) n(x,a)_t dx \\
	$$
\end{linenomath*}

\noindent The first term estimates the number of individuals recruiting from a one or two-year seed bank to a plant of size $x'$ and ant state $a'$ based on the recruit size distribution $\eta(x')$ and the probability of seedling survival ($\omega$) from germination (late summer) to the census (May).
This term is multiplied by $\rho_{0}(a')$, which gives the probability that a new recruit has ant state $a'$ at its first appearance in our census ($\sum_{a'}^{A}\rho_{0}(a')=1$). 
The second term represents all possible transitions from size $x$ and ant $a$ to size $x'$ and ant $a'$, conditioned on survival. 
Survival from initial size $x$ ($S(x,a;\pmb{\tau^S_{a}})$) and growth from size $x$ to $x'$ ($G(x',x,a;\pmb{\tau^G_{a}})$) are both dependent on initial size and ant state. 
As above, these functions include inter-annual variability through year-specific deviates that can vary by ant state ($\pmb{\tau_{a}}$). 
Ant transition function $\rho(a',a,x;\pmb{\tau^{\rho}})$ gives the probability that an individual transitions from ant state $a$ to $a'$ in the next census, conditional on initial size $x$. 
This function includes inter-annual variability through year-specific intercepts which are consistent across ant states ($\pmb{\tau^\rho}$).

\subsection*{Statistical modeling and parameter estimation}
We parameterized the IPM using a series of generalized linear mixed models (GLMMs) in a hierarchical Bayesian framework to serve as vital rate sub-models. 
Vital rate models included spatial and temporal random effects associated with plot and year variation, respectively, and included plant size ($log(cm^3)$; $x,x'$), ant partner state ($a,a'$), or both as fixed-effect predictor variables. 
In addition to vital rate models describing plant demographic performance, we also fit a sub-model to predict transition between ant states conditional on plant size and previous ant state. 
All models were fit using Stan and Rstan (\tom{}{Cite Stan and RStan.}). 
Unless otherwise mentioned, all models use vague priors.

@@

\paragraph{Seeds Per Flower}
With data from \tom{Miller 2007}{CITE}********, we fit a Bayesian generalized linear model for the number of seeds produced by every flower on a cholla ($\kappa(a')$) in year $t+1$ based on the ant partner $a'$ in year $t+1$, using a Negative Binomial distribution:
$$y^{\kappa} \sim 0 Truncated Negative Binomial(\hat{\kappa},\hat{\phi})$$
$$ \hat{\kappa } = \beta_{0}^{\kappa} \times a'$$
$$\hat{\phi} = \beta_{0}^{\phi}$$
Ant partners are included as predictors here because they reduce floral abortion rates and therefore may lead to higher numbers of seeds. 
\tom{}{This paragraph needs to address how we assigned seed number to the ``other'' category.}
